Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-15 paper/paper.tex
updated intro

Kiselyov \cite{kiselyov:sharing} presents a method for implementing eDSLs in
finally tagless form that generate a directed acyclic graph (DAG) with sharing.
However, as we will explain in sections~\ref{limithashcons} and
\ref{limitexplicit}, for DSL functions that return multiple outputs (e.g.,
tuples, lists, etc.), Kiselyov's method of implicitly detecting sharing may
require computation exponential in the size of the program, and his method of
explicitly declaring sharing is inapplicable.

In the toy example
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
its two outputs when writing its implementation that cannot be solved using
the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
issue in Section \ref{limitexplicit}. In our work, this translated into the
inability to process large library functions.

>>>>

Embedded DSL's have proven useful for many applications, and there are multiple
ways of doing the embedding. Domain experts are more comfortable with embeddings
presented as a collection of pure functions. On the other hand, optimizing code
generators and other downstream uses would be much easier to implement in the
context of a monad. In particular code graph generation outside of a monad does
not benefit from observable sharing in Haskell. Kiselyov \cite{kiselyov:sharing}
solves this problem by presenting a method for implementing eDSLs in finally
tagless form that generates a directed acyclic graph (DAG) with sharing.
However, as we will explain in sections~\ref{limithashcons} and
\ref{limitexplicit}, for DSL functions that return multiple outputs (e.g.,
tuples, lists, etc.), Kiselyov's method of implicitly detecting sharing may
require computation exponential in the size of the program, and his method of
explicitly declaring sharing is inapplicable.

In the toy example
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
the function \mintinline{haskell}{novel} will exhibit this issue. Since it
returns multiple outputs via a native tuple or list, it will diverge state used
to generate the DAG that cannot be captured by Kiselyov's explicit sharing
method. We will illustrate this implementation issue in Section
\ref{limitexplicit}. In our work, this translated into the inability to process
large library functions.
